---
title: "LLM and GAM Code"
author: "Jared Martin"
format: pdf
editor: visual
---

```{r}

library(tidyverse)
library(phenofit)
library(sf)
library(patchwork)
library(lme4)
library(lmerTest)
library(MuMIn)
library(performance)
library(caret)
library(mgcv)
library(xgboost)
library(ggtext)

```

```{r}

read_rds("../../../data/processed_trees_ndvi_lc_climate.rds") -> tree_ldc_metrics_climate 

```

```{r}

tree_ldc_metrics_climate %>%
  group_by(species_common, leaf_persistence) %>%
  summarize(n = length(unique(UID))) %>% 
  filter(n > 50) %>% 
  pull(1) -> species_list

pheno_vars <- c("Greenup_doy", "Maturity_doy", 
                "Senescence_doy", "Dormancy_doy")

# Define climate variables (same as before)
climate_vars <- list(
  Greenup_doy = c("late_winter_temp", "late_winter_prcp", "late_winter_mean_tmax","late_winter_prcp_days", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m"),
  Maturity_doy = c("late_spring_temp", "spring_prcp", "late_spring_mean_tmax","late_spring_prcp_days", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m"), 
  Senescence_doy = c("late_summer_prcp", "late_summer_prcp_days", "early_fall_mean_tmax", "early_fall_temp", "late_summer_mean_tmax", "late_summer_temp", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m"),
  Dormancy_doy = c("fall_cumulative_prcp", "fall_prcp_days", "fall_mean_tmax", "late_fall_temp", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m")
)

```


# Mixed Effect Model top level regressions for each species, unique intercpes for each tree. Using k-fold cross validation to check model preforamance.

```{r}

# Function to test model assumptions
test_assumptions <- function(model) {
  # Get residuals and fitted values
  resid <- residuals(model)
  fitted_vals <- fitted(model)
  
  # 1. Normality of residuals (Shapiro-Wilk test)
  # Note: Shapiro test limited to n <= 5000
  if (length(resid) <= 5000) {
    shapiro_test <- shapiro.test(resid)
    normality_p <- shapiro_test$p.value
  } else {
    # Use Kolmogorov-Smirnov for large samples
    ks_test <- ks.test(resid, "pnorm", mean(resid), sd(resid))
    normality_p <- ks_test$p.value
  }
  
  # 2. Homoscedasticity (Breusch-Pagan test)
  # Test if variance of residuals is constant
  # We'll use a simple test: correlation between |residuals| and fitted values
  bp_cor <- cor.test(abs(resid), fitted_vals)
  homoscedasticity_p <- bp_cor$p.value
  
  # 3. Normality of random effects
  ranef_vals <- unlist(ranef(model))
  if (length(ranef_vals) > 3 && length(ranef_vals) <= 5000) {
    ranef_shapiro <- shapiro.test(ranef_vals)
    ranef_normality_p <- ranef_shapiro$p.value
  } else if (length(ranef_vals) > 5000) {
    ranef_ks <- ks.test(ranef_vals, "pnorm", mean(ranef_vals), sd(ranef_vals))
    ranef_normality_p <- ranef_ks$p.value
  } else {
    ranef_normality_p <- NA
  }
  
  # 4. Check for outliers (standardized residuals > 3)
  std_resid <- resid / sd(resid)
  n_outliers <- sum(abs(std_resid) > 3)
  prop_outliers <- n_outliers / length(resid)
  
  return(list(
    normality_p = normality_p,
    homoscedasticity_p = homoscedasticity_p,
    ranef_normality_p = ranef_normality_p,
    n_outliers = n_outliers,
    prop_outliers = prop_outliers,
    pass_normality = normality_p > 0.05,
    pass_homoscedasticity = homoscedasticity_p > 0.05,
    pass_ranef_normality = ifelse(is.na(ranef_normality_p), TRUE, ranef_normality_p > 0.05)
  ))
}

# K-fold CV function (same as before)
cv_lmer_both <- function(formula_str, data, k = 10, response_var) {
  set.seed(123)
  folds <- createFolds(1:nrow(data), k = k, returnTrain = FALSE)
  
  predictions_conditional <- numeric(nrow(data))
  predictions_marginal <- numeric(nrow(data))
  
  for(fold_idx in 1:length(folds)) {
    test_idx <- folds[[fold_idx]]
    train_data <- data[-test_idx, ]
    test_data <- data[test_idx, ]
    
    model <- tryCatch({
      lmer(as.formula(formula_str), 
           data = train_data, 
           REML = FALSE)
    }, error = function(e) {
      return(NULL)
    })
    
    if (!is.null(model)) {
      predictions_conditional[test_idx] <- predict(model, newdata = test_data,
                                                   allow.new.levels = TRUE)
      predictions_marginal[test_idx] <- predict(model, newdata = test_data, 
                                                re.form = NA)
    } else {
      predictions_conditional[test_idx] <- NA
      predictions_marginal[test_idx] <- NA
    }
  }
  
  actual <- data[[response_var]]
  valid_idx <- !is.na(predictions_conditional) & !is.na(predictions_marginal) & !is.na(actual)
  
  if (sum(valid_idx) > 0) {
    cv_r2_conditional <- cor(actual[valid_idx], predictions_conditional[valid_idx])^2
    rmse_conditional <- sqrt(mean((actual[valid_idx] - predictions_conditional[valid_idx])^2))
    
    cv_r2_marginal <- cor(actual[valid_idx], predictions_marginal[valid_idx])^2
    rmse_marginal <- sqrt(mean((actual[valid_idx] - predictions_marginal[valid_idx])^2))
  } else {
    cv_r2_conditional <- cv_r2_marginal <- NA
    rmse_conditional <- rmse_marginal <- NA
  }
  
  return(list(
    cv_r2_conditional = cv_r2_conditional,
    cv_r2_marginal = cv_r2_marginal,
    rmse_conditional = rmse_conditional,
    rmse_marginal = rmse_marginal
  ))
}

# Store results
model_results <- list()

for (i in pheno_vars) {
  for (j in species_list) {
    
    cat("\nFitting", i, "for", j, "...\n")
    
    # Prepare data for this species
    data <- tree_ldc_metrics_climate %>% 
      filter(species_common == j, year <= 2024) %>% 
      mutate(
        late_winter_temp_z = scale(late_winter_temp)[,1],
        late_winter_prcp_z = scale(late_winter_prcp)[,1],
        late_spring_mean_tmax_z = scale(late_spring_mean_tmax)[,1],
        spring_prcp_z = scale(spring_prcp)[,1],
        late_summer_prcp_z = scale(late_summer_prcp)[,1],
        early_fall_mean_tmax_z = scale(early_fall_mean_tmax)[,1],
        fall_cumulative_prcp_z = scale(fall_cumulative_prcp)[,1],
        fall_mean_tmax_z = scale(fall_mean_tmax)[,1],
        late_spring_cumulative_prcp_z = scale(late_spring_cumulative_prcp)[,1],
        spring_temp_z = scale(spring_temp)[,1],
        irrigated_90m_z = scale(irrigated_90m)[,1],
        summer_temp_z = scale(summer_temp)[,1],
        late_summer_prcp_z = scale(late_summer_prcp)[,1]
      )
    
    if (nrow(data) < 20 || sum(!is.na(data[[i]])) < 20) {
      cat("  Skipping - insufficient data\n")
      next
    }
    
    predictors <- climate_vars[[i]]
    formula_str <- paste(i, "~", paste(predictors, collapse = " + "), "+ (1 | UID)")
    
    # Fit full model
    full_model <- tryCatch({
      lmer(as.formula(formula_str), 
           data = data, 
           REML = TRUE)
    }, error = function(e) {
      message(paste("  Error fitting model:", e$message))
      return(NULL)
    })
    
    # Run k-fold CV
    cv_results <- tryCatch({
      cv_lmer_both(formula_str, data, k = 10, response_var = i)
    }, error = function(e) {
      message(paste("  Error in CV:", e$message))
      return(NULL)
    })
    
    # Test assumptions
    assumptions <- tryCatch({
      test_assumptions(full_model)
    }, error = function(e) {
      message(paste("  Error testing assumptions:", e$message))
      return(NULL)
    })
    
    # Store results
    if (!is.null(full_model) && !is.null(cv_results) && !is.null(assumptions)) {
      library(MuMIn)
      r2_insample <- r.squaredGLMM(full_model)
      
      model_results[[paste(j, i, sep = "_")]] <- list(
        species = j,
        phenology = i,
        n_obs = nrow(data),
        n_trees = n_distinct(data$UID),
        model = full_model,
        summary = summary(full_model),
        r2m_insample = r2_insample[1, "R2m"],
        r2c_insample = r2_insample[1, "R2c"],
        cv_r2_marginal = cv_results$cv_r2_marginal,
        cv_r2_conditional = cv_results$cv_r2_conditional,
        cv_rmse_marginal = cv_results$rmse_marginal,
        cv_rmse_conditional = cv_results$rmse_conditional,
        # Assumption test results
        normality_p = assumptions$normality_p,
        homoscedasticity_p = assumptions$homoscedasticity_p,
        ranef_normality_p = assumptions$ranef_normality_p,
        n_outliers = assumptions$n_outliers,
        prop_outliers = assumptions$prop_outliers,
        pass_normality = assumptions$pass_normality,
        pass_homoscedasticity = assumptions$pass_homoscedasticity,
        pass_ranef_normality = assumptions$pass_ranef_normality
      )
      
      cat("  CV R²m:", round(cv_results$cv_r2_marginal, 3), 
          "| R²c:", round(cv_results$cv_r2_conditional, 3), "\n")
      cat("  Assumptions: Normality p =", round(assumptions$normality_p, 3),
          "| Homoscedasticity p =", round(assumptions$homoscedasticity_p, 3), "\n")
    }
  }
}

```

## Summarize and Results

### R2, RMSE and Testing model assumptions for all the regression models

```{r}

results_table <- map_df(model_results, function(x) {
  data.frame(
    species = x$species,
    phenology = x$phenology,
    n_obs = x$n_obs,
    n_trees = x$n_trees,
    # R² metrics
    r2m_insample = x$r2m_insample,
    r2c_insample = x$r2c_insample,
    cv_r2_marginal = x$cv_r2_marginal,
    cv_r2_conditional = x$cv_r2_conditional,
    # RMSE
    cv_rmse_marginal = x$cv_rmse_marginal,
    cv_rmse_conditional = x$cv_rmse_conditional,
    # Random effect contribution
    random_effect_contribution = x$cv_r2_conditional - x$cv_r2_marginal,
    # Assumption tests
    normality_p = x$normality_p,
    homoscedasticity_p = x$homoscedasticity_p,
    ranef_normality_p = x$ranef_normality_p,
    n_outliers = x$n_outliers,
    prop_outliers = x$prop_outliers,
    # Pass/fail flags
    pass_normality = x$pass_normality,
    pass_homoscedasticity = x$pass_homoscedasticity,
    pass_ranef_normality = x$pass_ranef_normality,
    # Overall assessment
    all_assumptions_met = x$pass_normality & x$pass_homoscedasticity & x$pass_ranef_normality,
    row.names = NULL
  )
})

results_table

```

### Breakdown by Table by Species or phenological metric

```{r}

# R² comparison - faceted by phenology
results_table %>%
  select(species, phenology, cv_r2_marginal, cv_r2_conditional) %>%
  mutate(random_effect_contribution = cv_r2_conditional - cv_r2_marginal) %>%
  pivot_longer(cols = starts_with("cv_r2"), 
               names_to = "type", 
               values_to = "r2") %>%
  mutate(type = ifelse(type == "cv_r2_marginal", "Marginal (Climate only)", 
                       "Conditional (Climate + Tree ID)")) %>%
  ggplot(aes(x = reorder(species, random_effect_contribution), y = r2, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~phenology, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Marginal (Climate only)" = "coral", 
                                "Conditional (Climate + Tree ID)" = "steelblue")) +
  labs(title = "Cross-Validated R²: Climate vs Climate + Tree Identity",
       subtitle = "Species ordered by tree identity contribution (gap size)",
       x = "", y = "CV R²", fill = "") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 8))

# RMSE comparison - faceted by phenology
results_table %>%
  select(species, phenology, cv_rmse_marginal, cv_rmse_conditional) %>%
  pivot_longer(cols = starts_with("cv_rmse"), 
               names_to = "type", 
               values_to = "rmse") %>%
  mutate(type = ifelse(type == "cv_rmse_marginal", "Marginal", "Conditional")) %>%
  ggplot(aes(x = reorder(species, -rmse), y = rmse, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~phenology, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Marginal" = "coral", 
                                "Conditional" = "steelblue")) +
  labs(title = "Cross-Validated RMSE: Average Prediction Error",
       subtitle = "By phenology stage",
       x = "", y = "RMSE (days)", fill = "") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 8))

# R² comparison - faceted by species
results_table %>%
  select(species, phenology, cv_r2_marginal, cv_r2_conditional) %>%
  pivot_longer(cols = starts_with("cv_r2"), 
               names_to = "type", 
               values_to = "r2") %>%
  mutate(type = ifelse(type == "cv_r2_marginal", "Marginal (Climate only)", 
                       "Conditional (Climate + Tree ID)")) %>%
  ggplot(aes(x = phenology, y = r2, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~species, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Marginal (Climate only)" = "coral", 
                                "Conditional (Climate + Tree ID)" = "steelblue")) +
  labs(title = "Cross-Validated R²: Climate vs Climate + Tree Identity",
       subtitle = "By species",
       x = "", y = "CV R²", fill = "") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 9))

```

```{r}

# Safer version with error checking
check_model_visually <- function(species_name, pheno_var) {
  model_name <- paste(species_name, pheno_var, sep = "_")
  
  # Check if model exists
  if (!model_name %in% names(model_results)) {
    cat("Model not found:", model_name, "\n")
    cat("Available models:", head(names(model_results)), "\n")
    return(NULL)
  }
  
  model <- model_results[[model_name]]$model
  
  # Check if model is valid
  if (is.null(model)) {
    cat("Model is NULL for:", model_name, "\n")
    return(NULL)
  }
  
  # Get residuals and fitted values
  resid <- residuals(model)
  fitted_vals <- fitted(model)
  
  # Check if we have valid data
  if (length(resid) == 0 || all(is.na(resid))) {
    cat("No valid residuals for:", model_name, "\n")
    return(NULL)
  }
  
  par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
  
  # 1. Q-Q plot (normality)
  qqnorm(resid, main = paste(species_name, "-", pheno_var, "\nQ-Q Plot"),
         pch = 16, col = rgb(0,0,0,0.2))
  qqline(resid, col = "red", lwd = 2)
  
  # 2. Residuals vs Fitted (homoscedasticity)
  plot(fitted_vals, resid,
       main = "Residuals vs Fitted",
       xlab = "Fitted", ylab = "Residuals",
       pch = 16, col = rgb(0,0,0,0.2))
  abline(h = 0, col = "red", lwd = 2)
  lowess_fit <- lowess(fitted_vals, resid)
  lines(lowess_fit, col = "blue", lwd = 2)
  
  # 3. Scale-Location (homoscedasticity)
  plot(fitted_vals, sqrt(abs(resid)),
       main = "Scale-Location",
       xlab = "Fitted", ylab = "√|Residuals|",
       pch = 16, col = rgb(0,0,0,0.2))
  lowess_fit2 <- lowess(fitted_vals, sqrt(abs(resid)))
  lines(lowess_fit2, col = "blue", lwd = 2)
  
  # 4. Histogram of residuals
  hist(resid, breaks = 30, main = "Residual Distribution",
       xlab = "Residuals", col = "lightblue")
  curve(dnorm(x, mean = mean(resid), sd = sd(resid)) * length(resid) * diff(range(resid))/30,
        add = TRUE, col = "red", lwd = 2)
  
  par(mfrow = c(1, 1))
  
  # Print summary stats
  cat("\n=== Diagnostics for", species_name, "-", pheno_var, "===\n")
  cat("Sample size:", length(resid), "\n")
  cat("Residual range:", round(range(resid), 2), "\n")
  cat("Residual SD:", round(sd(resid), 2), "\n")
  cat("Shapiro p-value:", format(results_table %>% 
                                   filter(species == species_name, phenology == pheno_var) %>% 
                                   pull(normality_p), scientific = TRUE), "\n\n")
}

# Get worst 5 models
worst_5 <- results_table %>%
  arrange(normality_p) %>%
  head(5)

print(worst_5 %>% select(species, phenology, normality_p, cv_r2_conditional))

# Check them one by one
for(i in 1:nrow(worst_5)) {
  cat("\n\n========== Model", i, "of", nrow(worst_5), "==========\n")
  check_model_visually(worst_5$species[i], worst_5$phenology[i])
  
  if (i < nrow(worst_5)) {
    response <- readline(prompt = "Press [enter] for next plot, or 'q' to quit: ")
    if (tolower(response) == 'q') break
  }
}

```

# Generalized Additive Mixed Model

```{r}

# Function to fit GAMM with CV for one species
fit_gamm_with_cv <- function(species_name, pheno_var, climate_predictors, data) {
  
  # Prepare data
  gamm_data <- data %>%
    filter(species_common == species_name, year < 2025) %>%
    select(all_of(c(pheno_var, "UID", climate_predictors))) 
  
  # Check sample size
  if (nrow(gamm_data) < 20) {
    return(NULL)
  }
  
  # Build formula with smooth terms
smooth_terms <- paste0("s(", climate_predictors, ", k = 6)", collapse = " + ")
formula_str <- paste(pheno_var, "~", smooth_terms, "+ s(UID, bs = 're')")
  
  cat("\nFitting:", species_name, "-", pheno_var)
  cat("\nFormula:", formula_str, "\n")
  
  # Fit full model
  full_model <- bam(
    as.formula(formula_str),
    data = gamm_data,
    method = "REML"
  )
  
  # Cross-validation
  set.seed(123)
  folds <- createFolds(1:nrow(gamm_data), k = 10, returnTrain = FALSE)
  
  predictions <- numeric(nrow(gamm_data))
  
  for(fold_idx in 1:length(folds)) {
    test_idx <- folds[[fold_idx]]
    train_data <- gamm_data[-test_idx, ]
    test_data <- gamm_data[test_idx, ]
    
    # Fit model on training fold
    cv_model <- tryCatch({
      bam(as.formula(formula_str),
          data = train_data,
          method = "REML")
    }, error = function(e) {
      return(NULL)
    })
    
    if (!is.null(cv_model)) {
      predictions[test_idx] <- predict(cv_model, newdata = test_data)
    } else {
      predictions[test_idx] <- NA
    }
  }
  
  # Calculate CV metrics
  actual <- gamm_data[[pheno_var]]
  valid_idx <- !is.na(predictions) & !is.na(actual)
  
  if (sum(valid_idx) > 0) {
    cv_r2 <- cor(actual[valid_idx], predictions[valid_idx])^2
    cv_rmse <- sqrt(mean((actual[valid_idx] - predictions[valid_idx])^2))
  } else {
    cv_r2 <- cv_rmse <- NA
  }
  
  cat("CV R²:", round(cv_r2, 3), "| RMSE:", round(cv_rmse, 2), "\n")
  
  return(list(
    species = species_name,
    phenology = pheno_var,
    n_obs = nrow(gamm_data),
    n_trees = n_distinct(gamm_data$UID),
    model = full_model,
    cv_r2 = cv_r2,
    cv_rmse = cv_rmse,
    predictions = predictions,
    actual = actual
  ))
}

# Fit GAMM models
gamm_results <- list()

for (i in pheno_vars) {
  for (j in species_list) {
    
    predictors <- climate_vars[[i]]
    
    result <- tryCatch({
      fit_gamm_with_cv(j, i, predictors, tree_ldc_metrics_climate)
    }, error = function(e) {
      message(paste("Error:", e$message))
      return(NULL)
    })
    
    if (!is.null(result)) {
      gamm_results[[paste(j, i, sep = "_")]] <- result
    }
  }
}

# Extract results
gamm_table <- map_df(gamm_results, function(x) {
  data.frame(
    species = x$species,
    phenology = x$phenology,
    n_obs = x$n_obs,
    n_trees = x$n_trees,
    gamm_cv_r2 = x$cv_r2,
    gamm_cv_rmse = x$cv_rmse,
    row.names = NULL
  )
})

print(gamm_table)

```

## Function to plot smooth effects pheno-metric

```{r}

plot_gamm_smooths <- function(model_name, gamm_results) {
  
  if (!model_name %in% names(gamm_results)) {
    cat("Model not found:", model_name, "\n")
    return(NULL)
  }
  
  model <- gamm_results[[model_name]]$model
  species <- gamm_results[[model_name]]$species
  pheno <- gamm_results[[model_name]]$phenology
  
  # Get smooth terms, excluding random effect
  smooth_labels <- sapply(model$smooth, function(x) x$label)
  predictor_idx <- which(!grepl("UID", smooth_labels))
  n_predictors <- length(predictor_idx)
  
  # Set layout
  if (n_predictors == 2) {
    par(mfrow = c(1, 2))
  } else if (n_predictors <= 4) {
    par(mfrow = c(2, 2))
  } else {
    par(mfrow = c(2, 3))
  }
  
  # Plot smooths
  for (i in 1:n_predictors) {
    idx <- predictor_idx[i]
    predictor_name <- model$smooth[[idx]]$term
    
    plot(model, 
         select = idx, 
         shade = TRUE, 
         main = if(i == 1) paste(species, "-", pheno) else "",
         xlab = predictor_name, 
         ylab = "Effect on DOY",
         rug = TRUE)
  }
  
  par(mfrow = c(1, 1))
  
  # Print summary
  print(summary(model))
}

# Usage
plot_gamm_smooths("Silver Maple_Dormancy_doy", gamm_results)

```
