---
title: "Model_development"
format: html
editor: visual
---

```{r}

library(tidyverse)
library(phenofit)
library(sf)
library(patchwork)
library(lme4)
library(lmerTest)
library(MuMIn)
library(performance)
library(caret)
library(mgcv)
library(xgboost)
library(ggtext)

```

```{r}

read_rds("../../data/processed_trees_ndvi_lc_climate.rds") -> tree_ldc_metrics_climate 

```

```{r}

tree_ldc_metrics_climate %>%
  group_by(species_common, leaf_persistence) %>%
  summarize(n = length(unique(UID))) %>% 
  filter(n > 50) %>% 
  pull(1) -> species_list

pheno_vars <- c("Greenup_doy", "Maturity_doy", 
                "Senescence_doy", "Dormancy_doy")

# Define climate variables (same as before)
climate_vars <- list(
  Greenup_doy = c("late_winter_temp", "late_winter_prcp", "late_winter_mean_tmax","late_winter_prcp_days", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m"),
  Maturity_doy = c("late_spring_temp", "spring_prcp", "late_spring_mean_tmax","late_spring_prcp_days", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m"), 
  Senescence_doy = c("late_summer_prcp", "late_summer_prcp_days", "early_fall_mean_tmax", "early_fall_temp", "late_summer_mean_tmax", "late_summer_temp", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m"),
  Dormancy_doy = c("fall_cumulative_prcp", "fall_prcp_days", "fall_mean_tmax", "late_fall_temp", "impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m")
)

```

# Corrilation Matrixs by species and dependent variables

```{r}

pheno_vars <- c("DER.sos_doy", "DER.eos_doy", "DER.pos_doy")  

sos_climate <- c("late_winter_prcp_days", "late_winter_temp", "late_winter_prcp", 
                 "winter_temp", "winter_prcp", "winter_prcp_days", 
                 "early_spring_temp", "early_spring_prcp_days", "early_spring_prcp",
                 "spring_prcp_days", "spring_temp", "spring_prcp", "impervious_90m", "irrigated_90m")

eos_climate <- c("summer_temp", "summer_prcp", "summer_prcp_days","late_summer_prcp", 
                 "late_summer_prcp_days", "late_summer_temp", "fall_temp", 
                 "fall_prcp", "fall_prcp_days", "early_fall_prcp", 
                 "early_fall_prcp_days", "early_fall_temp",
                 "late_fall_prcp", "late_fall_prcp_days", "late_fall_temp", "impervious_90m", "irrigated_90m")

pos_climate <- c("late_spring_temp", "late_spring_prcp", "late_spring_prcp_days", "impervious_90m", "irrigated_90m") 

all_climate_vars <- unique(c(sos_climate, eos_climate, pos_climate))

calc_species_cor <- function(species_name, data){
  data %>% 
    filter(year < 2025, species_common == species_name) %>% 
    select(all_of(c(pheno_vars, all_climate_vars))) %>% 
    cor(use = "pairwise.complete.obs") %>% 
    as.data.frame() %>% 
    rownames_to_column("pheno_var") %>% 
    filter(pheno_var %in% pheno_vars) %>% 
    select(pheno_var, all_of(all_climate_vars)) %>% 
    pivot_longer(-pheno_var, names_to = "climate_var", values_to = "correlation") %>% 
    filter(
      (pheno_var == "DER.sos_doy" & climate_var %in% sos_climate) |
      (pheno_var == "DER.eos_doy" & climate_var %in% eos_climate) |
      (pheno_var == "DER.pos_doy" & climate_var %in% pos_climate) |
      (pheno_var == "DER.pos_doy" & climate_var %in% pos_climate) |
      (pheno_var == "DER.pos_doy" & climate_var %in% pos_climate) |
      (pheno_var == "DER.pos_doy" & climate_var %in% pos_climate) |
      (pheno_var == "DER.pos_doy" & climate_var %in% pos_climate) 
    ) %>% 
    rename(!!species_name := correlation)
}

# Calculate for all species
map(species_list, ~calc_species_cor(.x, tree_ldc_metrics_climate)) %>% 
  compact() %>%  # Remove NULL values
  reduce(full_join, by = c("pheno_var", "climate_var")) -> all_cors 

all_cors %>% 
  pivot_longer(cols = -c(pheno_var, climate_var), 
               names_to = "species", 
               values_to = "correlation") %>% 
  mutate(group = "Individual Species") %>% 
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(pheno_var, climate_var) %>% 
      summarise(correlation = mean(correlation, na.rm = TRUE), .groups = "drop") %>% 
      mutate(species = "Average", group = "Average")
  ) %>% 
  
  ggplot(aes(x = species, y = climate_var, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 2) +
  scale_fill_gradient2(low = "red", mid = "white", high = "darkgreen",
                       midpoint = 0, limits = c(-1, 1), 
                       name = "Correlation") +
  facet_grid(pheno_var ~ group, scales = "free", space = "free") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 7),
    axis.title = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(title = "Phenology-Climate Correlations Across Species")

```

```{r}

# Updated phenology variables to match your labels
pheno_vars <- c("Greenup_doy", 
                "Senescence_doy", "Dormancy_doy", "Maturity_doy")

# Use only existing climate variables
greenup_climate <- c(
  "winter_temp", "winter_mean_tmax", "winter_prcp", "winter_prcp_days",
  "late_winter_temp", "late_winter_mean_tmax", "late_winter_prcp", "late_winter_prcp_days",
  "early_spring_temp", "early_spring_mean_tmax", "early_spring_prcp", "early_spring_prcp_days",
  "later_winter_cumulative_prcp", "early_spring_cumulative_prcp"
)

maturity_climate <- c(
  "spring_temp", "spring_mean_tmax", "spring_prcp", "spring_prcp_days",
  "late_spring_temp", "late_spring_mean_tmax", "late_spring_prcp", "late_spring_prcp_days",
  "spring_prcp_cumulative_prcp", "late_spring_cumulative_prcp"
)

senescence_climate <- c(
  "summer_temp", "summer_mean_tmax", "summer_prcp", "summer_prcp_days",
  "late_summer_temp", "late_summer_mean_tmax", "late_summer_prcp", "late_summer_prcp_days",
  "early_fall_temp", "early_fall_mean_tmax", "early_fall_prcp", "early_fall_prcp_days",
  "summer_cumulative_prcp", "late_summer_cumulative_prcp", "early_fall_cumulative_prcp"
)

dormancy_climate <- c(
  "fall_temp", "fall_mean_tmax", "fall_prcp", "fall_prcp_days",
  "late_fall_temp", "late_fall_prcp", "late_fall_prcp_days",
  "fall_cumulative_prcp", "late_fall_cumulative_prcp"
)

all_climate_vars <- unique(c(greenup_climate, maturity_climate, senescence_climate, 
                             dormancy_climate))

calc_species_cor <- function(species_name, data){
  data %>% 
    filter(year < 2025, species_common == species_name) %>% 
    select(all_of(c(pheno_vars, all_climate_vars))) %>% 
    cor(use = "pairwise.complete.obs") %>% 
    as.data.frame() %>% 
    rownames_to_column("pheno_var") %>% 
    filter(pheno_var %in% pheno_vars) %>% 
    select(pheno_var, all_of(all_climate_vars)) %>% 
    pivot_longer(-pheno_var, names_to = "climate_var", values_to = "correlation") %>% 
    filter(
      (pheno_var == "Greenup_doy" & climate_var %in% greenup_climate) |
      (pheno_var == "Maturity_doy" & climate_var %in% maturity_climate) |
      (pheno_var == "Senescence_doy" & climate_var %in% senescence_climate) |
      (pheno_var == "Dormancy_doy" & climate_var %in% dormancy_climate) 
    ) %>% 
    rename(!!species_name := correlation)
}

# Calculate for all species
map(species_list, ~calc_species_cor(.x, tree_ldc_metrics_climate)) %>% 
  compact() %>%
  reduce(full_join, by = c("pheno_var", "climate_var")) -> all_cors

# Labels
pheno_labels <- c(
  "DER.pos_doy" = "POS",
  "Greenup_doy" = "Green-up",
  "Maturity_doy" = "Maturity",
  "Senescence_doy" = "Senescence",
  "Dormancy_doy" = "Dormancy",
  "growing_season" = "Growing Season"  
)

# Calculate mean absolute correlations for each phenological phase (excluding Total)
pheno_mean_abs_cors <- all_cors %>% 
  pivot_longer(cols = -c(pheno_var, climate_var), 
               names_to = "species", 
               values_to = "correlation") %>% 
  group_by(pheno_var) %>% 
  summarise(mean_abs_cor = mean(abs(correlation), na.rm = TRUE), .groups = "drop") %>% 
  mutate(pheno_var_label = pheno_labels[pheno_var],
         label_with_mean = sprintf("%s (μ = %.2f)", pheno_var_label, mean_abs_cor)) %>% 
  select(pheno_var, label_with_mean)

# Create named vector for labeling (add Total without mean)
pheno_labeller <- c(
  setNames(pheno_mean_abs_cors$label_with_mean, pheno_mean_abs_cors$pheno_var),
  "Total" = "Total"
)

# Plot
all_cors %>% 
  pivot_longer(cols = -c(pheno_var, climate_var), 
               names_to = "species", 
               values_to = "correlation") %>% 
  mutate(group = "Individual Species") %>% 
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(pheno_var, climate_var) %>% 
      summarise(correlation = mean(correlation, na.rm = TRUE), .groups = "drop") %>% 
      mutate(species = "Mean", group = "Mean")
  ) %>% 
  # Add overall species summary statistics
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(species) %>% 
      summarise(
        mean_abs_cor = mean(abs(correlation), na.rm = TRUE),
        sd_cor = sd(correlation, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      pivot_longer(cols = c(mean_abs_cor, sd_cor), 
                   names_to = "climate_var", 
                   values_to = "correlation") %>% 
      mutate(
        pheno_var = "Total",
        group = "Individual Species",
        climate_var = case_when(
          climate_var == "mean_abs_cor" ~ "Species absolute mean",
          climate_var == "sd_cor" ~ "Species SD"
        )
      )
  ) %>% 
  # Add overall stats for Mean column
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(pheno_var, climate_var) %>% 
      summarise(correlation = mean(correlation, na.rm = TRUE), .groups = "drop") %>% 
      ungroup() %>%
      summarise(
        mean_abs_cor = mean(abs(correlation), na.rm = TRUE),
        sd_cor = sd(correlation, na.rm = TRUE)
      ) %>% 
      pivot_longer(cols = everything(), 
                   names_to = "climate_var", 
                   values_to = "correlation") %>% 
      mutate(
        pheno_var = "Total",
        species = "Mean",
        group = "Mean",
        climate_var = case_when(
          climate_var == "mean_abs_cor" ~ "Species absolute mean",
          climate_var == "sd_cor" ~ "Species SD",
          TRUE ~ climate_var
        )
      )
  ) %>% 
  mutate(pheno_var = case_when(
    pheno_var == "Total" ~ pheno_var,
    TRUE ~ pheno_labels[pheno_var]
  )) %>%
  mutate(pheno_var = factor(pheno_var, 
                            levels = c("Green-up", "Maturity", "Senescence", "Dormancy", "Total"))) %>%
  ggplot(aes(x = species, y = climate_var, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 3.5) +
  scale_fill_gradient2(low = "red", mid = "white", high = "darkgreen",
                       midpoint = 0, limits = c(-1, 1), 
                       name = NULL) +
  facet_grid(pheno_var ~ group, scales = "free", space = "free",
             labeller = labeller(pheno_var = pheno_labeller)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_markdown(size = 11),
    axis.title = element_blank(),
    strip.text = element_text(face = "bold", size = 12)
  )

```

```{r}

tree_ldc_metrics_climate %>%
  filter(year < 2025) %>% 
  ggplot(aes(impervious_90m, Maturity_doy))+
  geom_point()+
  geom_smooth()

```

```{r}

# Updated phenology variables to match your labels
pheno_vars <- c("Greenup_doy", 
                "Senescence_doy", "Dormancy_doy", "Maturity_doy")

# Use only existing climate variables
greenup_climate <- c(
  "winter_temp", "winter_mean_tmax", "winter_prcp", "winter_prcp_days",
  "late_winter_temp", "late_winter_mean_tmax", "late_winter_prcp", "late_winter_prcp_days",
  "early_spring_temp", "early_spring_mean_tmax", "early_spring_prcp", "early_spring_prcp_days",
  "later_winter_cumulative_prcp", "early_spring_cumulative_prcp"
)

maturity_climate <- c(
  "spring_temp", "spring_mean_tmax", "spring_prcp", "spring_prcp_days",
  "late_spring_temp", "late_spring_mean_tmax", "late_spring_prcp", "late_spring_prcp_days",
  "spring_prcp_cumulative_prcp", "late_spring_cumulative_prcp"
)

senescence_climate <- c(
  "summer_temp", "summer_mean_tmax", "summer_prcp", "summer_prcp_days",
  "late_summer_temp", "late_summer_mean_tmax", "late_summer_prcp", "late_summer_prcp_days",
  "early_fall_temp", "early_fall_mean_tmax", "early_fall_prcp", "early_fall_prcp_days",
  "summer_cumulative_prcp", "late_summer_cumulative_prcp", "early_fall_cumulative_prcp"
)

dormancy_climate <- c(
  "fall_temp", "fall_mean_tmax", "fall_prcp", "fall_prcp_days",
  "late_fall_temp", "late_fall_prcp", "late_fall_prcp_days",
  "fall_cumulative_prcp", "late_fall_cumulative_prcp"
)

all_climate_vars <- unique(c(greenup_climate, maturity_climate, senescence_climate, 
                             dormancy_climate))

calc_species_cor <- function(species_name, data){
  data %>% 
    filter(year < 2025, species_common == species_name) %>% 
    select(all_of(c(pheno_vars, all_climate_vars))) %>% 
    cor(use = "pairwise.complete.obs") %>% 
    as.data.frame() %>% 
    rownames_to_column("pheno_var") %>% 
    filter(pheno_var %in% pheno_vars) %>% 
    select(pheno_var, all_of(all_climate_vars)) %>% 
    pivot_longer(-pheno_var, names_to = "climate_var", values_to = "correlation") %>% 
    filter(
      (pheno_var == "Greenup_doy" & climate_var %in% greenup_climate) |
      (pheno_var == "Maturity_doy" & climate_var %in% maturity_climate) |
      (pheno_var == "Senescence_doy" & climate_var %in% senescence_climate) |
      (pheno_var == "Dormancy_doy" & climate_var %in% dormancy_climate) 
    ) %>% 
    rename(!!species_name := correlation)
}

# Calculate for all species
map(species_list, ~calc_species_cor(.x, tree_ldc_metrics_climate)) %>% 
  compact() %>%
  reduce(full_join, by = c("pheno_var", "climate_var")) -> all_cors

# Labels
pheno_labels <- c(
  "DER.pos_doy" = "POS",
  "Greenup_doy" = "Green-up",
  "Maturity_doy" = "Maturity",
  "Senescence_doy" = "Senescence",
  "Dormancy_doy" = "Dormancy",
  "growing_season" = "Growing Season"  
)

# Calculate mean absolute correlations for each phenological phase (excluding Total)
pheno_mean_abs_cors <- all_cors %>% 
  pivot_longer(cols = -c(pheno_var, climate_var), 
               names_to = "species", 
               values_to = "correlation") %>% 
  group_by(pheno_var) %>% 
  summarise(mean_abs_cor = mean(abs(correlation), na.rm = TRUE), .groups = "drop") %>% 
  mutate(pheno_var_label = pheno_labels[pheno_var],
         label_with_mean = sprintf("%s (μ = %.2f)", pheno_var_label, mean_abs_cor)) %>% 
  select(pheno_var_label, label_with_mean)

# Create named vector for labeling - keys should be the transformed names
pheno_labeller <- c(
  setNames(pheno_mean_abs_cors$label_with_mean, pheno_mean_abs_cors$pheno_var_label),
  "Total" = "Total"
)

# Plot
all_cors %>% 
  pivot_longer(cols = -c(pheno_var, climate_var), 
               names_to = "species", 
               values_to = "correlation") %>% 
  mutate(group = "Individual Species") %>% 
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(pheno_var, climate_var) %>% 
      summarise(correlation = mean(correlation, na.rm = TRUE), .groups = "drop") %>% 
      mutate(species = "Mean", group = "Mean")
  ) %>% 
  # Add overall species summary statistics
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(species) %>% 
      summarise(
        mean_abs_cor = mean(abs(correlation), na.rm = TRUE),
        sd_cor = sd(correlation, na.rm = TRUE),
        .groups = "drop"
      ) %>% 
      pivot_longer(cols = c(mean_abs_cor, sd_cor), 
                   names_to = "climate_var", 
                   values_to = "correlation") %>% 
      mutate(
        pheno_var = "Total",
        group = "Individual Species",
        climate_var = case_when(
          climate_var == "mean_abs_cor" ~ "Species absolute mean",
          climate_var == "sd_cor" ~ "Species SD"
        )
      )
  ) %>% 
  # Add overall stats for Mean column
  bind_rows(
    all_cors %>% 
      pivot_longer(cols = -c(pheno_var, climate_var), 
                   names_to = "species", 
                   values_to = "correlation") %>% 
      group_by(pheno_var, climate_var) %>% 
      summarise(correlation = mean(correlation, na.rm = TRUE), .groups = "drop") %>% 
      ungroup() %>%
      summarise(
        mean_abs_cor = mean(abs(correlation), na.rm = TRUE),
        sd_cor = sd(correlation, na.rm = TRUE)
      ) %>% 
      pivot_longer(cols = everything(), 
                   names_to = "climate_var", 
                   values_to = "correlation") %>% 
      mutate(
        pheno_var = "Total",
        species = "Mean",
        group = "Mean",
        climate_var = case_when(
          climate_var == "mean_abs_cor" ~ "Species absolute mean",
          climate_var == "sd_cor" ~ "Species SD",
          TRUE ~ climate_var
        )
      )
  ) %>% 
  mutate(pheno_var = case_when(
    pheno_var == "Total" ~ pheno_var,
    TRUE ~ pheno_labels[pheno_var]
  )) %>%
  mutate(pheno_var = factor(pheno_var, 
                            levels = c("Green-up", "Maturity", "Senescence", "Dormancy", "Total"))) %>%
  ggplot(aes(x = species, y = climate_var, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 3.5) +
  scale_fill_gradient2(low = "red", mid = "white", high = "darkgreen",
                       midpoint = 0, limits = c(-1, 1), 
                       name = NULL) +
  facet_grid(pheno_var ~ group, scales = "free", space = "free",
             labeller = labeller(pheno_var = pheno_labeller)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_markdown(size = 11),
    axis.title = element_blank(),
    strip.text = element_text(face = "bold", size = 12)
  )

ggsave("../../images/phenology_climate_correlations.png", 
       width = 16, height = 17, units = "in", dpi = 300)

```


```{r}

source("../../R/plot_year.R")

plot_year(tree_ldc_metrics_climate, "2020", "10006", metrics = c("Inflection")) 
plot_year(tree_ldc_metrics_climate, "2020", "10006", metrics = c("Derivative")) 



# Plot with smoother

df <- data.frame(
  t = tree_ldc_metrics_climate$season_data[[10]]$t,
  y = tree_ldc_metrics_climate$season_data[[10]]$y,
  lat = tree_ldc_metrics_climate$y_lat[10],
  long = tree_ldc_metrics_climate$x_long[10],
  species_botanice = tree_ldc_metrics_climate$species_botanic[10],
  species_common = tree_ldc_metrics_climate$species_common[10]  # adjust column name as needed
)%>%
  filter(year(t) < 2025)

# Plot with smoother
ggplot(df, aes(x = t, y = y)) +
  geom_point(alpha = 0.6, size = 2, color = "black") +
  labs(
    title = paste(df$species_botanice[1], "-", df$species_common[1]),
    subtitle = paste("Lat: ", round(df$lat[1], 4), "° | Long: ", round(df$long[1], 4), "°"),
    x = "Date",
    y = "NDVI"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, "mm")
  ) 


  ggsave("../../images/multiseason_tree_ndvi_plot.png", width = 10, height = 6, dpi = 300)
    
    
```

# Variable Selection for the Mixed Effect Panel Model

```{r, warning=FALSE}

data_complete <- tree_ldc_metrics_climate %>% 
  dplyr::select(Greenup_date:Dormancy_date, DER.pos_doy, UID, winter_prcp:water_250m, growing_season) %>% 
  mutate(Greenup_date = as.numeric(Greenup_date),
         Maturity_date = as.numeric(Maturity_date),
         DER.pos_doy = as.numeric(DER.pos_doy),
         Senescence_date = as.numeric(Senescence_date),
         Dormancy_date = as.numeric(Dormancy_date)) %>%  
  filter(complete.cases(.))

global_greening_model <- lmer(
  Greenup_date ~ winter_prcp + winter_prcp_days + winter_temp + 
    winter_mean_tmax + late_winter_prcp + late_winter_prcp_days + 
    late_winter_temp + late_winter_mean_tmax + spring_prcp + 
    spring_prcp_days + spring_temp + spring_mean_tmax + 
    early_spring_prcp + early_spring_prcp_days + early_spring_temp + 
    early_spring_mean_tmax + impervious_90m + irrigated_90m + 
    later_winter_cumulative_prcp + spring_prcp_cumulative_prcp + 
    early_spring_cumulative_prcp + (1 | UID),
  data = data_complete,
  REML = FALSE,
  na.action = "na.fail"
)

global_maturity_model <- lmer(
  Maturity_date ~ spring_prcp + spring_temp + spring_mean_tmax +
    late_spring_prcp + late_spring_temp + late_spring_mean_tmax +
    summer_prcp + summer_temp + summer_mean_tmax +
    spring_prcp_cumulative_prcp + late_spring_cumulative_prcp +
    impervious_90m + irrigated_90m + water_250m +
    (1 | UID),
  data = data_complete,
  REML = FALSE,
  na.action = "na.fail"
)

global_pos_model <- lmer(
  DER.pos_doy ~ spring_prcp + spring_temp + spring_mean_tmax +
    late_spring_prcp + late_spring_temp + late_spring_mean_tmax +
    summer_prcp + summer_temp + summer_mean_tmax +
    spring_prcp_cumulative_prcp + late_spring_cumulative_prcp +
    impervious_90m + irrigated_90m + water_250m +
    (1 | UID),
  data = data_complete,
  REML = FALSE,
  na.action = "na.fail"
)

global_senescence_model <- lmer(
  Senescence_date ~ late_summer_prcp + 
    #late_summer_temp + 
    late_summer_mean_tmax +
    summer_cumulative_prcp + 
    late_summer_cumulative_prcp +
    fall_prcp + 
    #fall_temp + 
    #fall_mean_tmax +
    early_fall_prcp + 
    early_fall_temp + 
    early_fall_mean_tmax +
    impervious_90m + irrigated_90m + water_250m +
    (1 | UID),
  data = data_complete,
  REML = FALSE,
  na.action = "na.fail"
)

global_dormancy_model <- lmer(
  Dormancy_date ~ fall_prcp + fall_temp + fall_mean_tmax +
    late_fall_prcp + late_fall_temp + late_fall_mean_tmax +
    fall_cumulative_prcp + late_fall_cumulative_prcp +
    winter_prcp + winter_temp + winter_mean_tmax +
    impervious_90m + irrigated_90m + water_250m +
    (1 | UID),
  data = data_complete,
  REML = FALSE,
  na.action = "na.fail"
)

global_gs_model <- lmer(
  growing_season ~ spring_prcp + spring_temp + 
                   summer_prcp + summer_temp +
                   late_summer_prcp + late_summer_temp +
                   fall_prcp + fall_temp +
                   spring_prcp_cumulative_prcp +
                   late_spring_cumulative_prcp +
                   summer_cumulative_prcp +
                   late_summer_cumulative_prcp +
                   fall_cumulative_prcp +
                   impervious_90m + irrigated_90m + water_250m +
                   (1 | UID),
  data = data_complete,
  REML = FALSE,
  na.action = "na.fail"
)

```

## Stepwise to find the best combination of 4 variables

```{r, warning=FALSE}

# all_models_greening <- dredge(global_greening_model, m.max = 2)
# all_maturity_models <- dredge(global_maturity_model, m.max = 4)
# all_pos_models <- dredge(global_pos_model, m.max = 4)
# all_senescence_models <- dredge(global_senescence_model, m.max = 4)
# all_dormancy_models <- dredge(global_dormancy_model, m.max = 4)
#all_gs_models <- dredge(global_gs_model, m.max = 4)

```

## Overview of the top 10 Best Models

```{r}

#head(all_models_greening, 10)

```

## Finding the best model

```{r}

# best_greening <- get.models(all_models_greening, subset = 1)[[1]]
# best_maturity <- get.models(all_maturity_models, subset = 1)[[1]]
# best_pos <- get.models(all_pos_models, subset = 1)[[1]]
# best_senescence <- get.models(all_senescence_models, subset = 1)[[1]]
# best_dormancy <- get.models(all_dormancy_models, subset = 1)[[1]]
#best_season <- get.models(all_gs_models, subset = 1)[[1]]
# 
# summary(best_greening)
# summary(best_maturity)
# summary(best_pos)
# summary(best_senescence)
# summary(best_dormancy)
#summary(best_season)

```

# Mixed Effect Model top level regressions for each species, unique intercpes for each tree. Using k-fold cross validation to check model preforamance.

```{r}

# Function to test model assumptions
test_assumptions <- function(model) {
  # Get residuals and fitted values
  resid <- residuals(model)
  fitted_vals <- fitted(model)
  
  # 1. Normality of residuals (Shapiro-Wilk test)
  # Note: Shapiro test limited to n <= 5000
  if (length(resid) <= 5000) {
    shapiro_test <- shapiro.test(resid)
    normality_p <- shapiro_test$p.value
  } else {
    # Use Kolmogorov-Smirnov for large samples
    ks_test <- ks.test(resid, "pnorm", mean(resid), sd(resid))
    normality_p <- ks_test$p.value
  }
  
  # 2. Homoscedasticity (Breusch-Pagan test)
  # Test if variance of residuals is constant
  # We'll use a simple test: correlation between |residuals| and fitted values
  bp_cor <- cor.test(abs(resid), fitted_vals)
  homoscedasticity_p <- bp_cor$p.value
  
  # 3. Normality of random effects
  ranef_vals <- unlist(ranef(model))
  if (length(ranef_vals) > 3 && length(ranef_vals) <= 5000) {
    ranef_shapiro <- shapiro.test(ranef_vals)
    ranef_normality_p <- ranef_shapiro$p.value
  } else if (length(ranef_vals) > 5000) {
    ranef_ks <- ks.test(ranef_vals, "pnorm", mean(ranef_vals), sd(ranef_vals))
    ranef_normality_p <- ranef_ks$p.value
  } else {
    ranef_normality_p <- NA
  }
  
  # 4. Check for outliers (standardized residuals > 3)
  std_resid <- resid / sd(resid)
  n_outliers <- sum(abs(std_resid) > 3)
  prop_outliers <- n_outliers / length(resid)
  
  return(list(
    normality_p = normality_p,
    homoscedasticity_p = homoscedasticity_p,
    ranef_normality_p = ranef_normality_p,
    n_outliers = n_outliers,
    prop_outliers = prop_outliers,
    pass_normality = normality_p > 0.05,
    pass_homoscedasticity = homoscedasticity_p > 0.05,
    pass_ranef_normality = ifelse(is.na(ranef_normality_p), TRUE, ranef_normality_p > 0.05)
  ))
}

# K-fold CV function (same as before)
cv_lmer_both <- function(formula_str, data, k = 10, response_var) {
  set.seed(123)
  folds <- createFolds(1:nrow(data), k = k, returnTrain = FALSE)
  
  predictions_conditional <- numeric(nrow(data))
  predictions_marginal <- numeric(nrow(data))
  
  for(fold_idx in 1:length(folds)) {
    test_idx <- folds[[fold_idx]]
    train_data <- data[-test_idx, ]
    test_data <- data[test_idx, ]
    
    model <- tryCatch({
      lmer(as.formula(formula_str), 
           data = train_data, 
           REML = FALSE)
    }, error = function(e) {
      return(NULL)
    })
    
    if (!is.null(model)) {
      predictions_conditional[test_idx] <- predict(model, newdata = test_data,
                                                   allow.new.levels = TRUE)
      predictions_marginal[test_idx] <- predict(model, newdata = test_data, 
                                                re.form = NA)
    } else {
      predictions_conditional[test_idx] <- NA
      predictions_marginal[test_idx] <- NA
    }
  }
  
  actual <- data[[response_var]]
  valid_idx <- !is.na(predictions_conditional) & !is.na(predictions_marginal) & !is.na(actual)
  
  if (sum(valid_idx) > 0) {
    cv_r2_conditional <- cor(actual[valid_idx], predictions_conditional[valid_idx])^2
    rmse_conditional <- sqrt(mean((actual[valid_idx] - predictions_conditional[valid_idx])^2))
    
    cv_r2_marginal <- cor(actual[valid_idx], predictions_marginal[valid_idx])^2
    rmse_marginal <- sqrt(mean((actual[valid_idx] - predictions_marginal[valid_idx])^2))
  } else {
    cv_r2_conditional <- cv_r2_marginal <- NA
    rmse_conditional <- rmse_marginal <- NA
  }
  
  return(list(
    cv_r2_conditional = cv_r2_conditional,
    cv_r2_marginal = cv_r2_marginal,
    rmse_conditional = rmse_conditional,
    rmse_marginal = rmse_marginal
  ))
}

# Store results
model_results <- list()

for (i in pheno_vars) {
  for (j in species_list) {
    
    cat("\nFitting", i, "for", j, "...\n")
    
    # Prepare data for this species
    data <- tree_ldc_metrics_climate %>% 
      filter(species_common == j, year <= 2024) %>% 
      mutate(
        late_winter_temp_z = scale(late_winter_temp)[,1],
        late_winter_prcp_z = scale(late_winter_prcp)[,1],
        late_spring_mean_tmax_z = scale(late_spring_mean_tmax)[,1],
        spring_prcp_z = scale(spring_prcp)[,1],
        late_summer_prcp_z = scale(late_summer_prcp)[,1],
        early_fall_mean_tmax_z = scale(early_fall_mean_tmax)[,1],
        fall_cumulative_prcp_z = scale(fall_cumulative_prcp)[,1],
        fall_mean_tmax_z = scale(fall_mean_tmax)[,1],
        late_spring_cumulative_prcp_z = scale(late_spring_cumulative_prcp)[,1],
        spring_temp_z = scale(spring_temp)[,1],
        irrigated_90m_z = scale(irrigated_90m)[,1],
        summer_temp_z = scale(summer_temp)[,1],
        late_summer_prcp_z = scale(late_summer_prcp)[,1]
      )
    
    if (nrow(data) < 20 || sum(!is.na(data[[i]])) < 20) {
      cat("  Skipping - insufficient data\n")
      next
    }
    
    predictors <- climate_vars[[i]]
    formula_str <- paste(i, "~", paste(predictors, collapse = " + "), "+ (1 | UID)")
    
    # Fit full model
    full_model <- tryCatch({
      lmer(as.formula(formula_str), 
           data = data, 
           REML = TRUE)
    }, error = function(e) {
      message(paste("  Error fitting model:", e$message))
      return(NULL)
    })
    
    # Run k-fold CV
    cv_results <- tryCatch({
      cv_lmer_both(formula_str, data, k = 10, response_var = i)
    }, error = function(e) {
      message(paste("  Error in CV:", e$message))
      return(NULL)
    })
    
    # Test assumptions
    assumptions <- tryCatch({
      test_assumptions(full_model)
    }, error = function(e) {
      message(paste("  Error testing assumptions:", e$message))
      return(NULL)
    })
    
    # Store results
    if (!is.null(full_model) && !is.null(cv_results) && !is.null(assumptions)) {
      library(MuMIn)
      r2_insample <- r.squaredGLMM(full_model)
      
      model_results[[paste(j, i, sep = "_")]] <- list(
        species = j,
        phenology = i,
        n_obs = nrow(data),
        n_trees = n_distinct(data$UID),
        model = full_model,
        summary = summary(full_model),
        r2m_insample = r2_insample[1, "R2m"],
        r2c_insample = r2_insample[1, "R2c"],
        cv_r2_marginal = cv_results$cv_r2_marginal,
        cv_r2_conditional = cv_results$cv_r2_conditional,
        cv_rmse_marginal = cv_results$rmse_marginal,
        cv_rmse_conditional = cv_results$rmse_conditional,
        # Assumption test results
        normality_p = assumptions$normality_p,
        homoscedasticity_p = assumptions$homoscedasticity_p,
        ranef_normality_p = assumptions$ranef_normality_p,
        n_outliers = assumptions$n_outliers,
        prop_outliers = assumptions$prop_outliers,
        pass_normality = assumptions$pass_normality,
        pass_homoscedasticity = assumptions$pass_homoscedasticity,
        pass_ranef_normality = assumptions$pass_ranef_normality
      )
      
      cat("  CV R²m:", round(cv_results$cv_r2_marginal, 3), 
          "| R²c:", round(cv_results$cv_r2_conditional, 3), "\n")
      cat("  Assumptions: Normality p =", round(assumptions$normality_p, 3),
          "| Homoscedasticity p =", round(assumptions$homoscedasticity_p, 3), "\n")
    }
  }
}

```

## Summarize and Results

### R2, RMSE and Testing model assumptions for all the regression models

```{r}

results_table <- map_df(model_results, function(x) {
  data.frame(
    species = x$species,
    phenology = x$phenology,
    n_obs = x$n_obs,
    n_trees = x$n_trees,
    # R² metrics
    r2m_insample = x$r2m_insample,
    r2c_insample = x$r2c_insample,
    cv_r2_marginal = x$cv_r2_marginal,
    cv_r2_conditional = x$cv_r2_conditional,
    # RMSE
    cv_rmse_marginal = x$cv_rmse_marginal,
    cv_rmse_conditional = x$cv_rmse_conditional,
    # Random effect contribution
    random_effect_contribution = x$cv_r2_conditional - x$cv_r2_marginal,
    # Assumption tests
    normality_p = x$normality_p,
    homoscedasticity_p = x$homoscedasticity_p,
    ranef_normality_p = x$ranef_normality_p,
    n_outliers = x$n_outliers,
    prop_outliers = x$prop_outliers,
    # Pass/fail flags
    pass_normality = x$pass_normality,
    pass_homoscedasticity = x$pass_homoscedasticity,
    pass_ranef_normality = x$pass_ranef_normality,
    # Overall assessment
    all_assumptions_met = x$pass_normality & x$pass_homoscedasticity & x$pass_ranef_normality,
    row.names = NULL
  )
})

results_table

```

### Breakdown by Table by Species or phenological metric

```{r}

# R² comparison - faceted by phenology
results_table %>%
  select(species, phenology, cv_r2_marginal, cv_r2_conditional) %>%
  mutate(random_effect_contribution = cv_r2_conditional - cv_r2_marginal) %>%
  pivot_longer(cols = starts_with("cv_r2"), 
               names_to = "type", 
               values_to = "r2") %>%
  mutate(type = ifelse(type == "cv_r2_marginal", "Marginal (Climate only)", 
                       "Conditional (Climate + Tree ID)")) %>%
  ggplot(aes(x = reorder(species, random_effect_contribution), y = r2, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~phenology, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Marginal (Climate only)" = "coral", 
                                "Conditional (Climate + Tree ID)" = "steelblue")) +
  labs(title = "Cross-Validated R²: Climate vs Climate + Tree Identity",
       subtitle = "Species ordered by tree identity contribution (gap size)",
       x = "", y = "CV R²", fill = "") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 8))

# RMSE comparison - faceted by phenology
results_table %>%
  select(species, phenology, cv_rmse_marginal, cv_rmse_conditional) %>%
  pivot_longer(cols = starts_with("cv_rmse"), 
               names_to = "type", 
               values_to = "rmse") %>%
  mutate(type = ifelse(type == "cv_rmse_marginal", "Marginal", "Conditional")) %>%
  ggplot(aes(x = reorder(species, -rmse), y = rmse, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~phenology, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Marginal" = "coral", 
                                "Conditional" = "steelblue")) +
  labs(title = "Cross-Validated RMSE: Average Prediction Error",
       subtitle = "By phenology stage",
       x = "", y = "RMSE (days)", fill = "") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 8))

# R² comparison - faceted by species
results_table %>%
  select(species, phenology, cv_r2_marginal, cv_r2_conditional) %>%
  pivot_longer(cols = starts_with("cv_r2"), 
               names_to = "type", 
               values_to = "r2") %>%
  mutate(type = ifelse(type == "cv_r2_marginal", "Marginal (Climate only)", 
                       "Conditional (Climate + Tree ID)")) %>%
  ggplot(aes(x = phenology, y = r2, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~species, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Marginal (Climate only)" = "coral", 
                                "Conditional (Climate + Tree ID)" = "steelblue")) +
  labs(title = "Cross-Validated R²: Climate vs Climate + Tree Identity",
       subtitle = "By species",
       x = "", y = "CV R²", fill = "") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 9))

```

```{r}

# Safer version with error checking
check_model_visually <- function(species_name, pheno_var) {
  model_name <- paste(species_name, pheno_var, sep = "_")
  
  # Check if model exists
  if (!model_name %in% names(model_results)) {
    cat("Model not found:", model_name, "\n")
    cat("Available models:", head(names(model_results)), "\n")
    return(NULL)
  }
  
  model <- model_results[[model_name]]$model
  
  # Check if model is valid
  if (is.null(model)) {
    cat("Model is NULL for:", model_name, "\n")
    return(NULL)
  }
  
  # Get residuals and fitted values
  resid <- residuals(model)
  fitted_vals <- fitted(model)
  
  # Check if we have valid data
  if (length(resid) == 0 || all(is.na(resid))) {
    cat("No valid residuals for:", model_name, "\n")
    return(NULL)
  }
  
  par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
  
  # 1. Q-Q plot (normality)
  qqnorm(resid, main = paste(species_name, "-", pheno_var, "\nQ-Q Plot"),
         pch = 16, col = rgb(0,0,0,0.2))
  qqline(resid, col = "red", lwd = 2)
  
  # 2. Residuals vs Fitted (homoscedasticity)
  plot(fitted_vals, resid,
       main = "Residuals vs Fitted",
       xlab = "Fitted", ylab = "Residuals",
       pch = 16, col = rgb(0,0,0,0.2))
  abline(h = 0, col = "red", lwd = 2)
  lowess_fit <- lowess(fitted_vals, resid)
  lines(lowess_fit, col = "blue", lwd = 2)
  
  # 3. Scale-Location (homoscedasticity)
  plot(fitted_vals, sqrt(abs(resid)),
       main = "Scale-Location",
       xlab = "Fitted", ylab = "√|Residuals|",
       pch = 16, col = rgb(0,0,0,0.2))
  lowess_fit2 <- lowess(fitted_vals, sqrt(abs(resid)))
  lines(lowess_fit2, col = "blue", lwd = 2)
  
  # 4. Histogram of residuals
  hist(resid, breaks = 30, main = "Residual Distribution",
       xlab = "Residuals", col = "lightblue")
  curve(dnorm(x, mean = mean(resid), sd = sd(resid)) * length(resid) * diff(range(resid))/30,
        add = TRUE, col = "red", lwd = 2)
  
  par(mfrow = c(1, 1))
  
  # Print summary stats
  cat("\n=== Diagnostics for", species_name, "-", pheno_var, "===\n")
  cat("Sample size:", length(resid), "\n")
  cat("Residual range:", round(range(resid), 2), "\n")
  cat("Residual SD:", round(sd(resid), 2), "\n")
  cat("Shapiro p-value:", format(results_table %>% 
                                   filter(species == species_name, phenology == pheno_var) %>% 
                                   pull(normality_p), scientific = TRUE), "\n\n")
}

# Get worst 5 models
worst_5 <- results_table %>%
  arrange(normality_p) %>%
  head(5)

print(worst_5 %>% select(species, phenology, normality_p, cv_r2_conditional))

# Check them one by one
for(i in 1:nrow(worst_5)) {
  cat("\n\n========== Model", i, "of", nrow(worst_5), "==========\n")
  check_model_visually(worst_5$species[i], worst_5$phenology[i])
  
  if (i < nrow(worst_5)) {
    response <- readline(prompt = "Press [enter] for next plot, or 'q' to quit: ")
    if (tolower(response) == 'q') break
  }
}

```

# Generalized Additive Mixed Model

```{r}

# Function to fit GAMM with CV for one species
fit_gamm_with_cv <- function(species_name, pheno_var, climate_predictors, data) {
  
  # Prepare data
  gamm_data <- data %>%
    filter(species_common == species_name, year < 2025) %>%
    select(all_of(c(pheno_var, "UID", climate_predictors))) 
  
  # Check sample size
  if (nrow(gamm_data) < 20) {
    return(NULL)
  }
  
  # Build formula with smooth terms
smooth_terms <- paste0("s(", climate_predictors, ", k = 6)", collapse = " + ")
formula_str <- paste(pheno_var, "~", smooth_terms, "+ s(UID, bs = 're')")
  
  cat("\nFitting:", species_name, "-", pheno_var)
  cat("\nFormula:", formula_str, "\n")
  
  # Fit full model
  full_model <- bam(
    as.formula(formula_str),
    data = gamm_data,
    method = "REML"
  )
  
  # Cross-validation
  set.seed(123)
  folds <- createFolds(1:nrow(gamm_data), k = 10, returnTrain = FALSE)
  
  predictions <- numeric(nrow(gamm_data))
  
  for(fold_idx in 1:length(folds)) {
    test_idx <- folds[[fold_idx]]
    train_data <- gamm_data[-test_idx, ]
    test_data <- gamm_data[test_idx, ]
    
    # Fit model on training fold
    cv_model <- tryCatch({
      bam(as.formula(formula_str),
          data = train_data,
          method = "REML")
    }, error = function(e) {
      return(NULL)
    })
    
    if (!is.null(cv_model)) {
      predictions[test_idx] <- predict(cv_model, newdata = test_data)
    } else {
      predictions[test_idx] <- NA
    }
  }
  
  # Calculate CV metrics
  actual <- gamm_data[[pheno_var]]
  valid_idx <- !is.na(predictions) & !is.na(actual)
  
  if (sum(valid_idx) > 0) {
    cv_r2 <- cor(actual[valid_idx], predictions[valid_idx])^2
    cv_rmse <- sqrt(mean((actual[valid_idx] - predictions[valid_idx])^2))
  } else {
    cv_r2 <- cv_rmse <- NA
  }
  
  cat("CV R²:", round(cv_r2, 3), "| RMSE:", round(cv_rmse, 2), "\n")
  
  return(list(
    species = species_name,
    phenology = pheno_var,
    n_obs = nrow(gamm_data),
    n_trees = n_distinct(gamm_data$UID),
    model = full_model,
    cv_r2 = cv_r2,
    cv_rmse = cv_rmse,
    predictions = predictions,
    actual = actual
  ))
}

# Fit GAMM models
gamm_results <- list()

for (i in pheno_vars) {
  for (j in species_list) {
    
    predictors <- climate_vars[[i]]
    
    result <- tryCatch({
      fit_gamm_with_cv(j, i, predictors, tree_ldc_metrics_climate)
    }, error = function(e) {
      message(paste("Error:", e$message))
      return(NULL)
    })
    
    if (!is.null(result)) {
      gamm_results[[paste(j, i, sep = "_")]] <- result
    }
  }
}

# Extract results
gamm_table <- map_df(gamm_results, function(x) {
  data.frame(
    species = x$species,
    phenology = x$phenology,
    n_obs = x$n_obs,
    n_trees = x$n_trees,
    gamm_cv_r2 = x$cv_r2,
    gamm_cv_rmse = x$cv_rmse,
    row.names = NULL
  )
})

print(gamm_table)

```

## Function to plot smooth effects pheno-metric

```{r}

plot_gamm_smooths <- function(model_name, gamm_results) {
  
  if (!model_name %in% names(gamm_results)) {
    cat("Model not found:", model_name, "\n")
    return(NULL)
  }
  
  model <- gamm_results[[model_name]]$model
  species <- gamm_results[[model_name]]$species
  pheno <- gamm_results[[model_name]]$phenology
  
  # Get smooth terms, excluding random effect
  smooth_labels <- sapply(model$smooth, function(x) x$label)
  predictor_idx <- which(!grepl("UID", smooth_labels))
  n_predictors <- length(predictor_idx)
  
  # Set layout
  if (n_predictors == 2) {
    par(mfrow = c(1, 2))
  } else if (n_predictors <= 4) {
    par(mfrow = c(2, 2))
  } else {
    par(mfrow = c(2, 3))
  }
  
  # Plot smooths
  for (i in 1:n_predictors) {
    idx <- predictor_idx[i]
    predictor_name <- model$smooth[[idx]]$term
    
    plot(model, 
         select = idx, 
         shade = TRUE, 
         main = if(i == 1) paste(species, "-", pheno) else "",
         xlab = predictor_name, 
         ylab = "Effect on DOY",
         rug = TRUE)
  }
  
  par(mfrow = c(1, 1))
  
  # Print summary
  print(summary(model))
}

# Usage
plot_gamm_smooths("Silver Maple_Dormancy_doy", gamm_results)

```

# XGBoost (Gradient Boosted Machines)

```{r}

set.seed(1)

# Modified function that trains final model after CV
fit_xgboost_model <- function(species_name, pheno_var, climate_predictors, data) {
  
  # Prepare data
  xgb_data <- data %>%
    filter(species_common == species_name, year < 2025) %>%
    select(all_of(c(pheno_var, "UID", climate_predictors))) %>%
    na.omit()
  
  if (nrow(xgb_data) < 20) return(NULL)
  
  # Add tree-level features to capture hierarchy
  xgb_data <- xgb_data %>%
    group_by(UID) %>%
    mutate(
      tree_mean_pheno = mean(.data[[pheno_var]], na.rm = TRUE),
      tree_sd_pheno = sd(.data[[pheno_var]], na.rm = TRUE),
      tree_n_obs = n()
    ) %>%
    ungroup() %>%
    mutate(UID_numeric = as.numeric(as.factor(UID)))
  
  # Create feature matrix
  feature_cols <- c(climate_predictors, "tree_mean_pheno", "tree_n_obs", "UID_numeric")
  X <- as.matrix(xgb_data %>% select(all_of(feature_cols)))
  y <- xgb_data[[pheno_var]]
  
  # 10-fold CV for evaluation
  set.seed(123)
  folds <- createFolds(1:nrow(xgb_data), k = 10, returnTrain = FALSE)
  predictions <- numeric(nrow(xgb_data))
  actuals <- y
  
  for(fold_idx in 1:length(folds)) {
    test_idx <- folds[[fold_idx]]
    train_idx <- setdiff(1:nrow(xgb_data), test_idx)
    
    dtrain <- xgb.DMatrix(data = X[train_idx, ], label = y[train_idx])
    dtest <- xgb.DMatrix(data = X[test_idx, ], label = y[test_idx])
    
    xgb_model <- xgb.train(
      data = dtrain,
      nrounds = 500,
      objective = "reg:squarederror",
      max_depth = 4,
      eta = 0.01,
      subsample = 0.7,
      colsample_bytree = 0.7,
      verbose = 0
    )
    
    predictions[test_idx] <- predict(xgb_model, dtest)
  }
  
  # Calculate CV metrics
  errors <- actuals - predictions
  cv_r2 <- cor(actuals, predictions)^2
  cv_rmse <- sqrt(mean(errors^2))
  cv_mae <- mean(abs(errors))
  cv_mape <- mean(abs(errors / actuals)) * 100 
  
  cat("\n  Training final model on all data...")
  dtrain_all <- xgb.DMatrix(data = X, label = y)
  
  final_model <- xgb.train(
    data = dtrain_all,
    nrounds = 500,
    objective = "reg:squarederror",
    max_depth = 4,
    eta = 0.01,
    subsample = 0.7,
    colsample_bytree = 0.7,
    verbose = 0
  )
  
  # Get feature importance from final model
  importance <- xgb.importance(model = final_model)
  
  # Store UID mapping for prediction
  uid_mapping <- data.frame(
    UID = xgb_data$UID,
    UID_numeric = xgb_data$UID_numeric
  ) %>% distinct()
  
  # Store tree-level stats for prediction
  tree_stats <- xgb_data %>%
    group_by(UID) %>%
    summarise(
      tree_mean_pheno = mean(.data[[pheno_var]], na.rm = TRUE),
      tree_sd_pheno = sd(.data[[pheno_var]], na.rm = TRUE),
      tree_n_obs = n(),
      .groups = "drop"
    )
  
  return(list(
    # Model objects
    model = final_model,
    
    # Model metadata
    species = species_name,
    phenology = pheno_var,
    climate_predictors = climate_predictors,
    feature_cols = feature_cols,
    
    # Training data info
    n_obs = nrow(xgb_data),
    uid_mapping = uid_mapping,
    tree_stats = tree_stats,
    
    # CV performance metrics
    cv_r2 = cv_r2,
    cv_rmse = cv_rmse,
    cv_mae = cv_mae,
    cv_mape = cv_mape,
    
    # CV predictions for diagnostics
    cv_predictions = predictions,
    cv_actuals = actuals,
    cv_errors = errors,
    
    # Feature importance
    importance = importance
  ))
}

# Run for all models
xgb_results <- list()

for (i in pheno_vars) {
  for (j in species_list) {
    cat("\nFitting XGBoost:", i, "for", j, "...")
    
    result <- tryCatch({
      fit_xgboost_model(j, i, climate_vars[[i]], tree_ldc_metrics_climate)
    }, error = function(e) {
      cat(" ERROR:", e$message)
      return(NULL)
    })
    
    if (!is.null(result)) {
      xgb_results[[paste(j, i, sep = "_")]] <- result
      cat(" CV R²:", round(result$cv_r2, 3))
    }
  }
}

# Save all models
cat("\n\nSaving models...\n")
saveRDS(xgb_results, "xgb_models_full.rds")
cat("Models saved to: xgb_models_full.rds\n")

# Extract results table
xgb_table <- map_df(xgb_results, function(x) {
  data.frame(
    species = x$species,
    phenology = x$phenology,
    n_obs = x$n_obs,
    RMSE = x$cv_rmse,
    MAE = x$cv_mae,
    MAPE = x$cv_mape,
    R2 = x$cv_r2
  )
})

# Save feature importance for all models
importance_all <- map_df(names(xgb_results), function(model_name) {
  x <- xgb_results[[model_name]]
  x$importance %>%
    mutate(
      species = x$species,
      phenology = x$phenology
    )
})

```

## Summary Stats Tables

```{r}
data.frame(
  Phenological_Stage = names(climate_vars),
  Variables = sapply(climate_vars, function(x) paste(x, collapse = ", "))
)

xgb_table %>%
  group_by(phenology) %>%
  summarise(
    N_Species = n(),
    Total_Obs = sum(n_obs),
    Mean_R2 = mean(R2, na.rm = TRUE),
    Mean_RMSE = mean(RMSE, na.rm = TRUE),
    Mean_MAE = mean(MAE, na.rm = TRUE)
  ) %>%
  arrange(Mean_RMSE)

xgb_table %>%
  group_by(phenology, species) %>%
  summarise(
    Total_Obs = n(),  # or sum(n_obs) if n_obs is a column
    Mean_R2 = mean(R2, na.rm = TRUE),
    Mean_RMSE = mean(RMSE, na.rm = TRUE),
    Mean_MAE = mean(MAE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(phenology = factor(phenology, 
                            levels = c("Greenup_doy", "Maturity_doy", "Senescence_doy",
                                       "Dormancy_doy"))) %>%
  arrange(phenology, species) %>% 
  write_csv("xgb_prefromnce_table.csv")
  

importance_all %>% 
  select(-Cover, -Frequency) %>% 
  pivot_wider(names_from = Feature, values_from = Gain) %>% 
  mutate(across(where(is.numeric), ~round(.x, 3))) #%>%
  #write_csv("xgb_importance.csv")
  

```

## predicting Climate senarios

```{r}

xgb_models <- readRDS("xgb_models_full.rds")

cat("Loaded", length(xgb_models), "models\n")

predict_climate_scenario <- function(models_list, scenario_data, scenario_name = "Scenario") {
  
  results <- map_df(names(models_list), function(model_name) {
    
    model_obj <- models_list[[model_name]]
    
    cat("\nPredicting:", model_obj$species, "-", model_obj$phenology)
    
    # Filter data for this species
    species_data <- scenario_data %>%
      filter(species_common == model_obj$species)
    
    if (nrow(species_data) == 0) {
      cat(" - No data for this species")
      return(NULL)
    }
    
    # Add tree-level features from training data
    species_data <- species_data %>%
      left_join(model_obj$tree_stats, by = "UID", 
                suffix = c("", "_from_training")) %>%
      left_join(model_obj$uid_mapping, by = "UID",
                suffix = c("", "_from_mapping"))
    
    # Check if we have new UIDs not in training
    duplicate_cols <- grep("_from_training$|_from_mapping$", names(species_data), value = TRUE)
    if (length(duplicate_cols) > 0) {
      cat("\n  ⚠️  Warning: Duplicate columns detected:", paste(duplicate_cols, collapse = ", "))
      cat("\n  Keeping scenario values, not training values\n")
    }
    
    if (any(is.na(species_data$UID_numeric))) {
      # Assign new numeric IDs
      max_uid <- max(model_obj$uid_mapping$UID_numeric)
      species_data <- species_data %>%
        mutate(
          UID_numeric = ifelse(is.na(UID_numeric), 
                              max_uid + row_number(), 
                              UID_numeric)
        )
      
      # For new trees, use species average
      species_avg <- mean(model_obj$tree_stats$tree_mean_pheno, na.rm = TRUE)
      species_data <- species_data %>%
        mutate(
          tree_mean_pheno = ifelse(is.na(tree_mean_pheno), species_avg, tree_mean_pheno),
          tree_n_obs = ifelse(is.na(tree_n_obs), 1, tree_n_obs)
        )
    }
    
    # Check if all required features exist
    missing_features <- setdiff(model_obj$feature_cols, names(species_data))
    if (length(missing_features) > 0) {
      cat(" - Missing features:", paste(missing_features, collapse = ", "))
      return(NULL)
    }
    
    # Create feature matrix
    X_new <- as.matrix(species_data %>% select(all_of(model_obj$feature_cols)))
    
    # Predict
    predictions <- predict(model_obj$model, X_new)
    
    # Return results
    data.frame(
      scenario = scenario_name,
      UID = species_data$UID,
      species = model_obj$species,
      phenology = model_obj$phenology,
      predicted_doy = predictions
    )
  })
  
  return(results)
}

```

```{r}

xgb_models <- readRDS("xgb_models_full.rds")

# Get current climate averages
current_climate <- read_csv("../../data_raw/nooa_daily_weather_obs/climate_variables/climate_baseline.csv")

prcp_vars <- c("late_winter_prcp", "spring_prcp", "late_summer_prcp", "fall_cumulative_prcp")

temp_vars <- c("late_winter_temp", "late_winter_mean_tmax", "late_spring_temp", 
               "late_spring_mean_tmax","early_fall_temp", "early_fall_mean_tmax",
               "fall_mean_tmax","late_fall_temp" )

landcover_vars <- c("impervious_90m", "irrigated_90m", "unirrigated_90m", "tree_canopy_90m", "water_90m")

# Scenario 1: Baseline (current average)
scenario_baseline <- expand_grid(
  UID = unique(tree_ldc_metrics_climate$UID),
  current_climate
  ) %>%
  left_join(
    tree_ldc_metrics_climate %>%
      select(UID, species_common, all_of(landcover_vars)) %>%
      distinct(),
    by = "UID") 

current_warm <- scenario_baseline %>%
  mutate(across(all_of(temp_vars), ~.x + 1.1),
         impervious_90m = mean(impervious_90m, na.rm = TRUE),
         irrigated_90m = mean(irrigated_90m, na.rm = TRUE),
         unirrigated_90m = mean(unirrigated_90m),
         water_90m = mean(water_90m),
         tree_canopy_90m = mean(tree_canopy_90m)
         )

# +2°C (modify ALL temperature variables)
scenario_warm <- scenario_baseline %>%
  mutate(across(all_of(temp_vars), ~.x + 2),
         impervious_90m = mean(impervious_90m, na.rm = TRUE),
         irrigated_90m = mean(irrigated_90m, na.rm = TRUE),
         unirrigated_90m = mean(unirrigated_90m),
         water_90m = mean(water_90m),
         tree_canopy_90m = mean(tree_canopy_90m)
         )

# +2°C + 10% more precipitation
scenario_warm_wet <- scenario_warm %>%
  mutate(across(all_of(prcp_vars), ~.x * 1.07))

# +2°C - 10% less precipitation
scenario_warm_dry <- scenario_warm %>%
  mutate(across(all_of(prcp_vars), ~.x * 0.93))

# ===== Predict all scenarios =====
cat("\n\n===== Predicting Climate Scenarios =====\n")

predictions_baseline <- predict_climate_scenario(xgb_models, scenario_baseline, "Baseline")
predictions_actual <- predict_climate_scenario(xgb_models, current_warm, "Actual Warming +1.1C")
predictions_warm <- predict_climate_scenario(xgb_models, scenario_warm, "+2C")
predictions_warm_wet <- predict_climate_scenario(xgb_models, scenario_warm_wet, "+2C +7% precip")
predictions_warm_dry <- predict_climate_scenario(xgb_models, scenario_warm_dry, "+2C -7% precip")

# Combine all scenarios
all_scenario_predictions <- bind_rows(
  predictions_baseline,
  predictions_actual,
  predictions_warm,
  predictions_warm_wet,
  predictions_warm_dry
)

```

```{r}

all_scenario_predictions %>% 
  group_by(species, phenology, scenario) %>% 
  summarise(mean_doy = mean(predicted_doy),
            sd = sd(predicted_doy)) 
  
```

```{r}

all_scenario_predictions %>% 
  mutate(
    phenology = factor(phenology, 
                      levels = c("Dormancy_doy", "Senescence_doy", "Maturity_doy", "Greenup_doy"),
                      labels = c("Dormancy", "Senescence", "Maturity", "Greenup"))  # Clean labels here!
   ) %>%
  ggplot(aes(x = species, y = predicted_doy, fill = scenario)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~phenology, ncol = 1, scales = "free_y", strip.position = "right") +
  scale_y_continuous(n.breaks = 6) +
  labs(
    x = "",
    y = "Day of Year",
    fill = "Scenario"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(.5, "lines"),
    legend.background = element_rect(fill = "white", color = "grey80"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y.right = element_text(angle = 90, face = "bold", size = 11),
    strip.background = element_rect(fill = "grey95", color = "grey80"),
    panel.spacing = unit(0.5, "lines")
  ) 
  ggsave("../../images/climatechange_impact.png", width = 10, height = 10, dpi = 300)

```

```{r}
all_scenario_predictions %>% 
  group_by(species, phenology, scenario) %>% 
  summarise(mean = mean(predicted_doy),
            sd = sd(predicted_doy),
            .groups = "drop") 
```

```{r}
get_partial_dependence <- function(species_name, phenology_stage, pred_var, 
                                   xgb_models, data, climate_vars) {
  
  # Get model
  model_key <- paste0(species_name, "_", phenology_stage)
  model_result <- xgb_models[[model_key]]
  model <- model_result$model
  
  # Prepare data
  xgb_data <- data %>%
    filter(species_common == species_name) %>%
    select(all_of(c(phenology_stage, "UID", climate_vars[[phenology_stage]]))) %>%
    na.omit()
  
  xgb_data <- xgb_data %>%
    group_by(UID) %>%
    mutate(
      tree_mean_pheno = mean(.data[[phenology_stage]], na.rm = TRUE),
      tree_n_obs = n()
    ) %>%
    ungroup() %>%
    mutate(UID_numeric = as.numeric(as.factor(UID)))
  
  feature_cols <- model_result$feature_cols
  train_data <- as.data.frame(xgb_data %>% select(all_of(feature_cols)))
  
  # Find the tree whose tree_mean_pheno is closest to the overall mean
  mean_pheno <- mean(train_data$tree_mean_pheno, na.rm = TRUE)
  tree_info <- train_data %>%
    distinct(UID_numeric, tree_mean_pheno, tree_n_obs) %>%
    mutate(dist_from_mean = abs(tree_mean_pheno - mean_pheno)) %>%
    arrange(dist_from_mean)
  
  typical_tree <- tree_info[1, ]  # Get first row using base R
  
  # Temperature range
  var_range <- seq(min(train_data[[pred_var]], na.rm = TRUE),
                   max(train_data[[pred_var]], na.rm = TRUE),
                   length.out = 100)
  
  # Create prediction grid
  pred_grid <- data.frame(matrix(ncol = length(feature_cols), nrow = length(var_range)))
  names(pred_grid) <- feature_cols
  
  # Set all variables
  for(col in feature_cols) {
    if(col == pred_var) {
      pred_grid[[col]] <- var_range  # Vary only this variable
    } else if(col == "UID_numeric") {
      pred_grid[[col]] <- typical_tree$UID_numeric  # Use the typical tree's ID
    } else if(col == "tree_mean_pheno") {
      pred_grid[[col]] <- typical_tree$tree_mean_pheno  # Use the typical tree's mean
    } else if(col == "tree_n_obs") {
      pred_grid[[col]] <- typical_tree$tree_n_obs  # Use the typical tree's n_obs
    } else {
      pred_grid[[col]] <- mean(train_data[[col]], na.rm = TRUE)  # Mean for climate vars
    }
  }
  
  # Make predictions
  predictions <- predict(model, as.matrix(pred_grid))
  
  pd_result <- data.frame(
    x = var_range,
    yhat = predictions,
    species = species_name,
    phenology = phenology_stage,
    variable = pred_var
  )
  
  names(pd_result)[1] <- pred_var
  return(pd_result)
}
```

```{r}

species_list <- c("American Elm", "American Linden", "Common Hackberry", 
                  "Eastern Cottonwood", "Green Ash", "Honey Locust",
                  "Northern Red Oak", "Plains Cottonwood", "Siberian Elm",
                  "Silver Maple", "Tree of heaven", "Western Catalpa")

# Loop through all species and create PD data
pd_all_species <- map_df(species_list, function(sp) {
  tryCatch({
    get_partial_dependence(
      species_name = sp,
      phenology_stage = "Senescence_doy",
      pred_var = "late_summer_mean_tmax",
      xgb_models = xgb_results,
      data = tree_ldc_metrics_climate,
      climate_vars = climate_vars
    )
  }, error = function(e) {
    cat("\nError for", sp, ":", e$message)
    return(NULL)
  })
})

# Plot all species together with both color and linetype
ggplot(pd_all_species, aes(x = late_summer_mean_tmax, y = yhat, 
                            color = species, linetype = species)) +
  geom_smooth(size = 1) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", 
                                   "longdash", "twodash", "solid", "dashed", 
                                   "dotted", "dotdash", "longdash", "twodash")) +
  labs(
    x = "Late Summer Mean Tmax (°C)",
    y = "Predicted Senescence DOY",
    title = "Senescence Response to Temperature Across Species",
    color = "Species",
    linetype = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

```{r}

# Define climate variables and their corresponding phenology stages
climate_var_mapping <- data.frame(
  variable = c("late_winter_prcp", "late_winter_temp", 
                   "late_spring_prcp_days", 
                   "late_spring_mean_tmax",
                   "spring_prcp",
                   "late_summer_mean_tmax", "early_fall_mean_tmax", "late_summer_prcp",
                   "fall_prcp_days", "fall_mean_tmax"),
  phenology = c("Greenup_doy", "Greenup_doy", "Maturity_doy",
                "Maturity_doy", "Maturity_doy", 
                "Senescence_doy", "Senescence_doy", "Senescence_doy",
                "Dormancy_doy", "Dormancy_doy")
)

species_list <- c("American Elm", "American Linden", "Common Hackberry", 
                  "Eastern Cottonwood", "Green Ash", "Honey Locust",
                  "Northern Red Oak", "Plains Cottonwood", "Siberian Elm",
                  "Silver Maple", "Tree of heaven", "Western Catalpa")

# Loop through all combinations
pd_all_results <- map_df(1:nrow(climate_var_mapping), function(i) {
  
  var <- climate_var_mapping$variable[i]
  pheno <- climate_var_mapping$phenology[i]
  
  cat("\nProcessing", var, "for", pheno, "\n")
  
  # Loop through species
  map_df(species_list, function(sp) {
    tryCatch({
      pd <- get_partial_dependence(
        species_name = sp,
        phenology_stage = pheno,
        pred_var = var,
        xgb_models = xgb_results,
        data = tree_ldc_metrics_climate,
        climate_vars = climate_vars
      )
      
      # Rename first column to standardized name
      pd <- pd %>%
        rename(x_value = 1) %>%
        select(x_value, yhat, species, phenology, variable)
      
      return(pd)
      
    }, error = function(e) {
      cat("  Error for", sp, ":", e$message, "\n")
      return(NULL)
    })
  })
})

# Create labels with units
var_labels <- c(
  "late_winter_prcp" = "Late Winter Precip (mm)",
  "late_winter_temp" = "Late Winter Temp (°C)",
  "late_spring_prcp_days" = "Late Spring Precip Days",
  "spring_temp" = "Spring Temp (°C)",
  "late_spring_mean_tmax" = "Late Spring Tmax (°C)",
  "spring_prcp" = "Spring Precip (mm)",
  "late_summer_mean_tmax" = "Late Summer Tmax (°C)",
  "early_fall_mean_tmax" = "Early Fall Tmax (°C)",
  "late_summer_prcp" = "Late Summer Precip (mm)",
  "fall_prcp_days" = "Fall Precip Days",
  "fall_mean_tmax" = "Fall Tmax (°C)",
  "fall_cumulative_prcp" = "Fall Cumulative Precip (mm)"
)

pd_all_results %>%
  mutate(
    pheno_label = case_when(
      phenology == "Greenup_doy" ~ "Green-up",
      phenology == "Maturity_doy" ~ "Maturity",
      phenology == "Senescence_doy" ~ "Senescence",
      phenology == "Dormancy_doy" ~ "Dormancy"
    ),
    # Identify variable type
    var_type = case_when(
      str_detect(variable, "prcp") ~ "Precipitation",
      str_detect(variable, "temp|tmax") ~ "Temperature"
    ),
    var_label = paste0(var_labels[variable], "\n(", pheno_label, ")"),
    # Order phenology stages
    pheno_order = factor(phenology, 
                        levels = c("Greenup_doy", "Maturity_doy", 
                                   "Senescence_doy", "Dormancy_doy"))
  ) %>%
  arrange(pheno_order, var_type, variable) %>%
  mutate(var_label = factor(var_label, levels = unique(var_label))) %>%
  ggplot(aes(x = x_value, y = yhat, color = species, linetype = species)) +
  geom_smooth(se = FALSE, size = 1) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", 
                                   "longdash", "twodash", "solid", "dashed", 
                                   "dotted", "dotdash", "longdash", "twodash")) +
  facet_wrap(~var_label, scales = "free", nrow = 6, ncol = 2) +
  labs(
    x = NULL, 
    y = "Predicted Phenological Timing (DOY)",
    color = "Species",
    linetype = "Species"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.5, "cm"),
    legend.key = element_blank(),
    strip.text = element_text(face = "bold", size = 9)
  )

ggsave("../../images/partial_depences_climnate_responces.png", 
       width = 16, height = 17, units = "in", dpi = 300)

```

```{r}

get_partial_dependence <- function(species_name, phenology_stage, pred_var, 
                                   xgb_models, data, climate_vars) {
  
  # Get model
  model_key <- paste0(species_name, "_", phenology_stage)
  model_result <- xgb_models[[model_key]]
  model <- model_result$model
  
  # Prepare data
  xgb_data <- data %>%
    filter(species_common == species_name) %>%
    select(all_of(c(phenology_stage, "UID", climate_vars[[phenology_stage]]))) %>%
    na.omit()
  
  xgb_data <- xgb_data %>%
    group_by(UID) %>%
    mutate(
      tree_mean_pheno = mean(.data[[phenology_stage]], na.rm = TRUE),
      tree_n_obs = n()
    ) %>%
    ungroup() %>%
    mutate(UID_numeric = as.numeric(as.factor(UID)))
  
  feature_cols <- model_result$feature_cols
  train_data <- as.data.frame(xgb_data %>% select(all_of(feature_cols)))
  
  # Find the tree whose tree_mean_pheno is closest to the overall mean
  mean_pheno <- mean(train_data$tree_mean_pheno, na.rm = TRUE)
  tree_info <- train_data %>%
    distinct(UID_numeric, tree_mean_pheno, tree_n_obs) %>%
    mutate(dist_from_mean = abs(tree_mean_pheno - mean_pheno)) %>%
    arrange(dist_from_mean)
  
  typical_tree <- tree_info[1, ]
  
  # Identify land cover variables
  land_cover_vars <- c("impervious_90m", "irrigated_90m", "tree_canopy_90m", 
                       "water_90m", "pervious_90m", "unirrigated_90m")
  land_cover_vars <- land_cover_vars[land_cover_vars %in% feature_cols]
  
  # Variable range
  var_range <- seq(min(train_data[[pred_var]], na.rm = TRUE),
                   max(train_data[[pred_var]], na.rm = TRUE),
                   length.out = 100)
  
  # Create prediction grid
  pred_grid <- data.frame(matrix(ncol = length(feature_cols), nrow = length(var_range)))
  names(pred_grid) <- feature_cols
  
  # Special handling for land cover variables
  if(pred_var %in% land_cover_vars) {
    
    # Get baseline proportions of other land cover types
    other_lc_vars <- setdiff(land_cover_vars, pred_var)
    baseline_lc <- sapply(other_lc_vars, function(v) mean(train_data[[v]], na.rm = TRUE))
    baseline_total <- sum(baseline_lc)
    
    # For each value of pred_var, adjust other land covers proportionally
    for(i in 1:length(var_range)) {
      pred_val <- var_range[i]
      remaining_proportion <- 1 - pred_val
      
      # Redistribute remaining proportion based on original ratios
      if(baseline_total > 0) {
        for(lc_var in other_lc_vars) {
          pred_grid[[lc_var]][i] <- baseline_lc[lc_var] * (remaining_proportion / baseline_total)
        }
      } else {
        # If baseline total is 0, distribute evenly
        for(lc_var in other_lc_vars) {
          pred_grid[[lc_var]][i] <- remaining_proportion / length(other_lc_vars)
        }
      }
      
      pred_grid[[pred_var]][i] <- pred_val
    }
    
  } else {
    # For non-land-cover variables, vary only the target variable
    pred_grid[[pred_var]] <- var_range
  }
  
  # Set all other variables (non-land-cover or those not yet set)
  for(col in feature_cols) {
    if(is.na(pred_grid[[col]][1])) {  # If not already set
      if(col == "UID_numeric") {
        pred_grid[[col]] <- typical_tree$UID_numeric
      } else if(col == "tree_mean_pheno") {
        pred_grid[[col]] <- typical_tree$tree_mean_pheno
      } else if(col == "tree_n_obs") {
        pred_grid[[col]] <- typical_tree$tree_n_obs
      } else {
        pred_grid[[col]] <- mean(train_data[[col]], na.rm = TRUE)
      }
    }
  }
  
  # Make predictions
  predictions <- predict(model, as.matrix(pred_grid))
  
  pd_result <- data.frame(
    x = var_range,
    yhat = predictions,
    species = species_name,
    phenology = phenology_stage,
    variable = pred_var
  )
  
  names(pd_result)[1] <- pred_var
  return(pd_result)
}

```

```{r}

# Define land cover variables (just impervious and irrigated)
land_cover_vars <- c("impervious_90m", "irrigated_90m")

# Create mapping for land cover across all stages
climate_var_mapping <- expand.grid(
  variable = land_cover_vars,
  phenology = c("Greenup_doy", "Maturity_doy", "Senescence_doy", "Dormancy_doy"),
  stringsAsFactors = FALSE
)

species_list <- c("American Elm", "American Linden", "Common Hackberry", 
                  "Eastern Cottonwood", "Green Ash", "Honey Locust",
                  "Northern Red Oak", "Plains Cottonwood", "Siberian Elm",
                  "Silver Maple", "Tree of heaven", "Western Catalpa")

# Loop through all combinations
pd_land_cover <- map_df(1:nrow(climate_var_mapping), function(i) {
  
  var <- climate_var_mapping$variable[i]
  pheno <- climate_var_mapping$phenology[i]
  
  cat("\nProcessing", var, "for", pheno, "\n")
  
  # Loop through species
  map_df(species_list, function(sp) {
    tryCatch({
      pd <- get_partial_dependence(
        species_name = sp,
        phenology_stage = pheno,
        pred_var = var,
        xgb_models = xgb_results,
        data = tree_ldc_metrics_climate,
        climate_vars = climate_vars
      )
      
      # Rename first column to standardized name
      pd <- pd %>%
        rename(x_value = 1) %>%
        select(x_value, yhat, species, phenology, variable)
      
      return(pd)
      
    }, error = function(e) {
      cat("  Error for", sp, ":", e$message, "\n")
      return(NULL)
    })
  })
})

# Create labels
var_labels <- c(
  "impervious_90m" = "Impervious Surface (%)",
  "irrigated_90m" = "Irrigated Land (%)"
)

# Plot land cover effects
pd_land_cover %>%
  mutate(
    pheno_label = case_when(
      phenology == "Greenup_doy" ~ "Green-up",
      phenology == "Maturity_doy" ~ "Maturity",
      phenology == "Senescence_doy" ~ "Senescence",
      phenology == "Dormancy_doy" ~ "Dormancy"
    ),
    pheno_order = factor(phenology, 
                        levels = c("Greenup_doy", "Maturity_doy", 
                                   "Senescence_doy", "Dormancy_doy")),
    var_label = var_labels[variable]
  ) %>%
  arrange(pheno_order, variable) %>%
  mutate(
    pheno_label = factor(pheno_label, levels = unique(pheno_label)),
    var_label = factor(var_label, levels = var_labels)
  ) %>%
  ggplot(aes(x = x_value, y = yhat, color = species)) +
  geom_smooth(se = FALSE, size = 1.2, alpha = 0.8) +
  scale_color_brewer(palette = "Paired") +
  facet_grid(pheno_label ~ var_label, scales = "free") +
  labs(
    x = "Land Cover (%)",
    y = "Predicted Phenological Timing (DOY)",
    title = "Land Cover Effects on Phenology"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.4, "cm"),
    legend.key = element_blank(),
    strip.text = element_text(face = "bold", size = 10)
  ) +
  guides(color = guide_legend(nrow = 3, ncol = 4, byrow = TRUE))

# Save
ggsave("../../images/pd_land_cover_impervious_irrigated.png", width = 10, height = 10, dpi = 300)

```

```{r}

# Plot land cover effects with ONE overall trend line
pd_land_cover %>%
  mutate(
    pheno_label = case_when(
      phenology == "Greenup_doy" ~ "Green-up",
      phenology == "Maturity_doy" ~ "Maturity",
      phenology == "Senescence_doy" ~ "Senescence",
      phenology == "Dormancy_doy" ~ "Dormancy"
    ),
    pheno_order = factor(phenology, 
                        levels = c("Greenup_doy", "Maturity_doy", 
                                   "Senescence_doy", "Dormancy_doy")),
    var_label = var_labels[variable]
  ) %>%
  arrange(pheno_order, variable) %>%
  mutate(
    pheno_label = factor(pheno_label, levels = unique(pheno_label)),
    var_label = factor(var_label, levels = var_labels)
  ) %>%
  ggplot(aes(x = x_value, y = yhat)) +
  # Individual species lines
  geom_smooth(aes(color = species), se = FALSE, size = 0.8, alpha = 0.5) +
  # ONE overall linear trend line (dashed, black)
  geom_smooth(method = "lm", se = FALSE, color = "black", 
              size = 1.2, linetype = "dashed") +
  scale_color_brewer(palette = "Paired") +
  facet_grid(pheno_label ~ var_label, scales = "free") +
  labs(
    x = "Land Cover (%)",
    y = "Predicted Phenological Timing (DOY)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.4, "cm"),
    legend.key = element_blank(),
    strip.text = element_text(face = "bold", size = 10)
  ) +
  guides(color = guide_legend(nrow = 3, ncol = 4, byrow = TRUE))

ggsave("../../images/pd_land_cover_impervious_irrigated.png", 
       width = 10, height = 10, dpi = 300)

```
